\documentclass[conference]{IEEEtran}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{hyperref}

% ── Theorem environments ─────────────────────────────────────────────
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{remark}[theorem]{Remark}

% ── Macros ───────────────────────────────────────────────────────────
\newcommand{\gadv}{g_{\mathrm{adv}}}
\newcommand{\gh}{g_h}
\newcommand{\clip}{\mathrm{clip}}
\newcommand{\MDP}{\mathcal{M}_{\mathrm{DP}}}
\newcommand{\Var}{\mathrm{Var}}

\title{DPAmplify: Byzantine Gradient Amplification\\via Differential-Privacy Structure}
\author{Matteo et al.}

\begin{document}
\maketitle

% ─────────────────────────────────────────────────────────────────────
\begin{abstract}
We present DPAmplify, an attack strategy in which a Byzantine client
in a differentially-private federated learning system exploits the
Gaussian DP mechanism to steer global model updates toward an arbitrary
target direction.  By crafting adversarial gradients whose $\ell_2$
norm equals the clipping threshold exactly, the attacker ensures that
the DP mechanism introduces zero bias and that its contributions
accumulate coherently across rounds.  We derive both a closed-form
\emph{upper bound} and an \emph{exact} signal-to-noise ratio formula
that fully accounts for honest gradient heterogeneity, providing a
sharper characterization of the conditions under which the attack is
most effective.
\end{abstract}

% ─────────────────────────────────────────────────────────────────────
\section{Introduction}

Federated learning (FL) enables collaborative model training without
sharing raw data.  Differentially-private variants (DP-FedAvg) add
Gaussian noise to clipped client gradients to bound information
leakage.  While DP noise is designed to protect privacy, its
\emph{structured} additive form opens an unexpected attack surface:
a Byzantine client can craft updates whose expected value after the
DP mechanism is fully aligned with a chosen target direction.

% ─────────────────────────────────────────────────────────────────────
\section{Background}

\paragraph{Gaussian DP mechanism.}
Given a gradient vector $g \in \mathbb{R}^d$, clipping threshold
$C>0$, and noise scale $\sigma>0$, the mechanism is
\[
  \MDP(g) = \clip(g, C) + \xi,\quad \xi \sim \mathcal{N}(0, \sigma^2 I),
\]
where $\clip(g, C) = g \cdot \min(1, C/\|g\|_2)$.

\paragraph{FedAvg aggregation.}
In each round, the server averages the noisy gradients of all $n$
clients:
\[
  \bar{g} = \frac{1}{n}\sum_{i=1}^{n} \MDP(g_i).
\]

% ─────────────────────────────────────────────────────────────────────
\section{The DPAmplify Attack}

\paragraph{Core observation.}
Let $\hat{v}$ be a unit target direction.  Set the adversarial
gradient $\gadv = C\,\hat{v}$.  Since $\|\gadv\|_2 = C$, clipping
is the identity:
\[
  \clip(\gadv, C) = \gadv,
  \quad\Rightarrow\quad
  \mathbb{E}[\MDP(\gadv)] = \gadv = C\,\hat{v}.
\]
The adversary's contribution to $\bar{g}$ is \emph{unattenuated in
expectation} and aligned with $\hat{v}$ regardless of the DP noise
level~$\sigma$.

\paragraph{Evasion property.}
Because all gradients are clipped to $\ell_2$ norm $\le C$ before
noise is added, the post-DP norm distributions of adversarial and
honest gradients are statistically indistinguishable, making
norm-based anomaly detectors ineffective.

% ─────────────────────────────────────────────────────────────────────
\section{Signal-to-Noise Analysis}
\label{sec:snr}

We characterise the attack strength through the signal-to-noise ratio
(SNR) of the aggregate gradient in the target direction.

% ── include the two-theorem block ────────────────────────────────────
\input{../theory/proofs/theorem_snr}

% ─────────────────────────────────────────────────────────────────────

\begin{corollary}[Simpler upper bound]
\label{cor:upper}
\[
  \mathrm{SNR} \;\le\; \frac{k\,C}{\sigma\sqrt{n-k}}.
\]
\end{corollary}

The interplay between Theorem~\ref{thm:snr-tight} and
Corollary~\ref{cor:upper} is central to our contribution:

\begin{quote}
\textit{Theorem~\ref{thm:snr-tight} (tight bound) reduces to the simpler
upper bound of Corollary~\ref{cor:upper} when honest gradient variance
is negligible relative to DP noise, i.e., $\Var_h \ll \sigma^2$.
Concretely, when $\Var_h = 0$, the tight bound becomes
$kC/(\sigma\sqrt{n})$, which differs from the upper bound only by the
factor $\sqrt{n/(n-k)} \approx 1$ for $k \ll n$.}
\end{quote}

Crucially, this characterises the \textbf{exact conditions} under which
the attack is most effective:
\begin{itemize}
  \item \textbf{IID / concentrated honest data} ($\Var_h \approx 0$):
        the two bounds coincide; the attack achieves near-maximum SNR.
        This is the canonical setting of prior DP-FL security analyses.
  \item \textbf{Non-IID / heterogeneous honest data} ($\Var_h \gg 0$):
        the tight bound predicts substantially lower SNR than the upper
        bound.  The attack is \emph{self-limiting} — honest gradient
        diversity acts as a natural defence.
\end{itemize}
This distinction is a stronger contribution than a single closed-form
bound: we identify the \emph{regime boundary} at $\Var_h \approx \sigma^2/n$
that separates the attack's effective and ineffective operating conditions.

% ─────────────────────────────────────────────────────────────────────
\section{Empirical Validation}

We validate both bounds with a NumPy proof-of-concept
(\texttt{poc.py}) using parameters $d=100$, $n=20$, $k=3$, $C=1.0$,
$\sigma=0.1$ over 300 simulated FedAvg rounds.

\begin{table}[h]
\centering
\begin{tabular}{lcc}
\hline
Quantity & Value & Error \\
\hline
SNR upper bound (Corollary~\ref{cor:upper}) & 7.28 & $35.6\%$ \\
SNR tight bound (Theorem~\ref{thm:snr-tight}) & $\approx 4.9$ & $<5\%$ \\
Empirical SNR & $\approx 4.7$ & — \\
\hline
\end{tabular}
\caption{SNR comparison ($T=300$ rounds, seed 42).  Estimated
$\widehat{\Var}_h \approx 0.010$ from post-clip honest projections.}
\end{table}

The tight bound matches simulation to within 5\%, while the upper bound
over-predicts by 36\% — confirming that $\Var_h$ is non-negligible in
the heterogeneous-data regime of this experiment.

% ─────────────────────────────────────────────────────────────────────
\section{Conclusion}

DPAmplify demonstrates that the Gaussian DP mechanism can be exploited
by a Byzantine client to coherently steer the global model update.  We
provide (i) a zero-bias construction, (ii) an evasion argument based
on norm indistinguishability, and (iii) both an upper-bound and a tight
SNR formula that together characterise when the attack succeeds and
when honest gradient heterogeneity naturally limits its effectiveness.

\end{document}
